Multimodal Language Learning in the Face of Missing Modalities

### 2023-01-17 Paper/main.tex
Before converting TMLR to SIGIR format

\newcommand{\CitationNeeded}[1]{{\textbf {\color{red}Cite #1}}}
\newcommand{\ProofRead}[1]{{\textbf {\color{red}Proofread please #1}}}
\newcommand{\Complete}[0]{{\textbf {\color{red}Complete it }}}
\newcommand{\Rephrase}[1]{{\textbf {\color{red}Rephrase please #1}}}
\newcommand{\TD}[1]{{\color{red}{\textbf{TODO}}} #1}
\newcommand{\OK}[1]{{\color{green}{\textbf{DONE}}} #1}
\newcommand{\ours}{\textsc{EMMA}}
% \newcommand{\ours}{\textsc{EMMA}}
\newcommand{\geom}{\textsc{Geometric Alignment}}
\newcommand{\supcon}{\textsc{SupCon}}

@@

    We propose \textit{extended multimodal alignment (EMMA)}, a generalized geometric method combined with a cross-entropy loss function that can be used to learn retrieval models that incorporate an arbitrary number of views of a particular piece of data, compounded by the challenge of retrieval when a modality becomes unavailable. Our study is motivated by needs in robotics and computer-human interaction, where an agent has many sensors and thus modalities with which a human may interact, both to communicate a desired goal and for the agent to recognize a desired target object. For such problems, there has been little research on integrating more than two modalities. While there have been widely popular works on self-supervised contrastive learning based on cross-entropy, there is an entirely separate family of approaches based on explicit geometric alignment. However, to the best of our knowledge there has been no work on combining the two approaches for multimodal learning. We propose to combine both families of approaches and argue that the two are complementary. We demonstrate the usability of our model on a grounded language object retrieval scenario, where an intelligent agent has to select an object given an unconstrained language command. We leverage four modalities including vision, depth sensing, text, and speech, and we show that our model converges approximately 5 times faster than previous strong baselines, and out-performs or is strongly competitive with state-of-the-art contrastive learning. % (in 80\% less time).
    %when modalities are ablated. 
    The code is publicly available on GitHub and will be included for the camera-ready version (it is redacted for anonymity). 
\end{abstract}

@@

The ultimate aim of this work is to take arbitrary input modalities about novel objects, including both spoken and written language, and build a model that allows a system to correctly retrieve that object given a subset of those input modalities. In this work, we propose our new method \ours{}, a generalized geometric method combined with a cross-entropy loss function that can be used to learn retrieval models that incorporate an arbitrary number of views of a particular piece of data. We demonstrate steps towards this ultimate goal with a mechanism that learns quickly from a variety of input signals and is robust to ablation of certain inputs. This is a generalization of approaches to grounded language learning in which specific input modalities are labeled with language to allow for future identification, but explicitly seeks to be agnostic about the nature of the individual sensory and linguistic inputs.

@@

In general there are two different families of approaches to multimodal learning in the literature. One family is based on classification and cross-entropy losses such as~\citet{NEURIPS2020_supervised_contrastive, chen2020simple} and the other family is geometric based such as~\citet{ICML22GeometricMultimodal,Carvalho-cooking-triplet,salvador2017cooking,triplet_loss_2021_CVPR}.
In this paper we propose to marry the two families and show that combining geometric approach with cross-entropy method is superior.

@@

@@

@@

Our proposed multimodal method is composed of two complementary parts. The first part is a geometric loss based on distances in the latent space, and the second part is a contrastive loss based on cross-entropy. The geometric loss is faster to learn while the cross-entropy method is more aligned with the downstream task of object retrieval. Hence we propose to combine them.

@@

To address this issue, we use pairwise distance optimization for all data points. %
Our method can be used for an arbitrary number of modalities. %
In contrast to our work, previous works that use triplet loss~\citep{GoLD_UMBC,triplet_loss_2021_CVPR} concatenate RGB and depth to form a single ``vision'' vector to handle three modalities, so they cannot robustly handle RGB or depth sensor ablation during test.
Moreover, our method has the advantage that it does not require providing hard negative examples.
Therefore, we modify the concept of triplet loss as follows. During training, we sample two different instances and their corresponding representations from all modalities into two sets---one positive set (referring to a specific object) and one negative set (referring to some different object) as shown in \cref{fig:emma-loss-trimmed}.
Unlike some prior triplet loss methods~\citep{GoLD_UMBC,triplet_loss_2021_CVPR}, the anchor is not randomly chosen from different modalities in each batch. Instead, in our setting, every item in the positive set becomes anchor once, and we minimize the distance between that item and other items in the positive set, while minimizing the distance between that item and all items in the negative set. It can be seen as an one-to-many relationship instead of an one-to-two relationship in the triplet loss formulation.
% the anchor and positive items are from the same object class, but in different modalities and each of the positives modalities becomes anchor once. 
We consider the following terminology:
% \todokdinline{I feel like using the terminology of triplet loss especially \textbf{anchor} is confusing here or maybe it needs to be rephrased. We can alternatively say that we consider all pairwise distances.}
% \todoffinline{I don't see the problem with anchor, but if you'd rather use a different term, what about query? As a note, while you do consider all pairwise distances but you treat some differently than others.}
\begin{itemize}
    % \item data point: a row in the csv/tsv file.
    \item Positive (Instance): A set of embeddings of one data point (e.g., an RGB image of an apple, corresponding depth image, text description, and speech description of the same apple) as shown with green rectangles in \cref{fig:emma-loss-trimmed}
    \item Negative (Instance): A set of embeddings of another data point of a different object (e.g., an RGB image of a mug, corresponding depth image, text description, and speech description of the same mug) as shown with orange rectangles in \cref{fig:emma-loss-trimmed}
    \item Anchor (Modality): Every modality of the positive set is chosen as the learning anchor once. In our formulation, anchor is a ``concept'' referring to the points in the positive sets rather than a single instance in itself. In \cref{fig:emma-loss-trimmed}, each of the 4 modalities is selected as an anchor once.
    %For our experiments, text is used. 
    The anchor is used as the basis for learning distances between positive and negative samples.
    % variance of text is lower and clusters of different objects are more distinct compared to other modalities, therefore, it can train a better model.
    % where we don't minimize the distance between the pair of embeddings from all modalities of the same data point, and we don't maximize the distance between positive and negative datapoints of the same modality.
    % \todo{explain this better}
    % \todokd{I rephrased it. Good?}
\end{itemize}

@@

\paragraph{Combining Geometric and Cross-Entropy Losses}
\label{subsec:emma}

@@

% 
% \begin{comment}
% 
% \begin{equation}\label{eq:emma}
%     % \mathcal{L} = \geom{} + \supcon{}
%     \ours{} = \geom{} + \supcon{}
% \end{equation}
% \end{comment}
% 
Compared to pure \supcon{}, combining these loss functions results in faster convergence and slightly improved performance when all modalities are available, and maintains improved performance when modalities are missing. Experimental results are presented in detail in \cref{sec:results}.

@@

@@

\subsection{Discussion}
Our proposed model performs well and learns fast, has been demonstrated to handle four modalities of shared information effectively, and is robust to test-time situations where information from one or more modalities is missing. There remains room for improvement. Specifically, the speech modality is harder to handle. \Cref{fig:3d-tsne} shows that although the relative position of instances are correct in the speech space, the distinction and clustering of different objects are not as good as the other three modalities.

>>>>

\newcommand{\CitationNeeded}[1]{{\textbf {\color{red}Cite #1}}}
\newcommand{\ProofRead}[1]{{\textbf {\color{red}Proofread please #1}}}
\newcommand{\Complete}[0]{{\textbf {\color{red}Complete it }}}
\newcommand{\Rephrase}[1]{{\textbf {\color{red}Rephrase please #1}}}
\newcommand{\TD}[1]{{\color{red}{\textbf{TODO}}} #1}
\newcommand{\OK}[1]{{\color{green}{\textbf{DONE}}} #1}
\newcommand{\DO}[1]{{\color{blue}{\textbf{PROG}}} #1}
\newcommand{\ours}{\textsc{EMMA}}
% \newcommand{\ours}{\textsc{EMMA}}
\newcommand{\geom}{\textsc{Geometric Alignment}}
\newcommand{\supcon}{\textsc{SupCon}}

@@

    Our study is motivated by needs in robotics and computer-human interaction, where an agent has many sensors and thus modalities with which a human may interact, both to communicate a desired goal and for the agent to recognize a desired target object. For such problems, there has been little research on integrating more than two modalities. While there have been widely popular works based on cross-entropy, there is an entirely separate family of approaches based on explicit geometric alignment known as contrastive learning. However, to the best of our knowledge there has been no work on combining the two approaches for multimodal learning. We propose to combine both families of approaches and argue that the two are complementary. We introduce \textit{extended multimodal alignment (EMMA)}, a geometric (distance-based) method combined with a contrastive loss function. Our method can be used to learn retrieval models that incorporate an arbitrary number of views of a particular piece of data, compounded by the challenge of retrieval when a modality becomes unavailable. We demonstrate the usability of our model on a grounded language object retrieval scenario, where an intelligent agent has to select an object given an unconstrained language command. We leverage four modalities including vision, depth sensing, text, and speech, and we show that our model converges approximately 5 times faster than previous strong baselines, and out-performs or is strongly competitive with state-of-the-art contrastive learning. % (in 80\% less time).
    %when modalities are ablated. 
    %TODO: another baseline where we use cross-entropy loss only is missing I guess.
    The code is publicly available on GitHub and will be included for the camera-ready version (it is redacted for anonymity). 
\end{abstract}

\section{TODO}
\begin{itemize}

    \item \TD modify it for SIGIR
    \item \TD EMMA as a contstrained optimization for Geometric portion to not satisfy it.
    \item \TD constrained SupCon per instance modality constraints: for this particular anchor and corresponding neighbors, we want this constraint to be satisfied.
    \item \OK Do you control for in batch negatives for supcon portion? Yes.
    \item \TD alternate training of supcon and emma. E.g. train supcon for 50 epochs and emma for the next 50 epochs
    \item \TD removing objects that have less than a number of instances? yes, but it's not a good story, since we want emma to work in low-data regime.
    \item \TD ablating number of data points and have a graph based on different number of training data --> maybe EMMA degrades less than SupCon and that would be good
    \item \TD use supcon portion to reweight the geometric values or use a method to have different scores for different pairs in the geometric loss. Targeting specific pair of embeddings.
    \item \TD it seems to be a tension between geometric portion and the supcon loss
    \item \OK another column for number of NAs in test and train and then correlation between them

    \item \OK add another column for the number of instances and then compute correlation between that and the differential scores. Take the correlation between number of instances and each column (metric) and then the correlation and the average of columns.

    \item listen to the speeches for NA descriptions and the ones where there is a significant difference in performance

    \item \OK listen to the speeches for empty transcriptions.

    \item \TD take a look at the depth images

    \item going forward we have two options: Improve the model, or do a deep analysis why our results are like this

    \item We could do something that addresses the "open questions" (which questions remain?)

    I guess the main open question is the mixed results and the fact that they are within the std dev of each other. Basically, I think they are unsure whether being faster is a good result or not.
    What the reader would take away?

    \item \DO What the rich analysis would consist of?

    \item Why are we getting what we are getting?

    Visualize individual data points and what are the failure modes.
    You can do the same for type by type (apples vs books)
    The percent that EMMA got right and SupCon got wrong and vice versa.

    have pictures, positive set and negative set,
    and all modalities and then show scores for all of them.

    \item \TD Different datasets or tasks
    \item \TD How can we tell when modalities are complementary?

    I think they meant when loss functions are complementary, am I wrong?

    \item \TD How can we tell what losses to use in what cases?

    \item  \TD Why didn't our approach work better, and what can we do to improve it?

    One reason could be that \supcon{} considers the distance between all pairs of data points in a batch.

    \item What's the next thing to try or add?

    We could add a weighting term to balance the effect of \supcon{} and \geom{}.

\end{itemize}

@@

The ultimate aim of this work is to take arbitrary input modalities about novel objects, including both spoken and written language, and build a model that allows a system to correctly retrieve that object given a subset of those input modalities. In this work, we propose our new method called \textit{extended multimodal alignment (\ours{})}, a generalized geometric method combined with a cross-entropy based supervised contrastive loss function~\cite{NEURIPS2020_supervised_contrastive}, which we abbreviate \supcon{}. Our method can be used to learn retrieval models that incorporate an arbitrary number of views of a particular piece of data. We demonstrate steps towards this ultimate goal with a mechanism that learns quickly from a variety of input signals and is robust to ablation of certain inputs. This is a generalization of approaches to grounded language learning in which specific input modalities are labeled with language to allow for future identification, but explicitly seeks to be agnostic about the nature of the individual sensory and linguistic inputs.

@@

In general there are two different families of approaches to multimodal learning in the literature. One family is based on classification and cross-entropy losses such as~\citet{NEURIPS2020_supervised_contrastive, chen2020simple} and the other family is geometric based such as~\citet{ICML22GeometricMultimodal,Carvalho-cooking-triplet,salvador2017cooking,triplet_loss_2021_CVPR}.
In this paper we propose to marry the two families and show that combining geometric approach with cross-entropy based \supcon{} loss function is superior.

@@

\paragraph{Core Concepts}
\label{sec:core-concepts}
There is a shared terminology across the methods we describe in this section, but they differ in what they incorporate. We explain the high-level concepts here and clarify their use cases for each method in the corresponding sections.
The three terms that are commonly used are \textit{anchor}, \textit{positive}, and \textit{negative}. Anchor refers to a data point that acts as a point of reference against which other data points are compared. Positive and negative refer to a data point or a set of data points that are similar and dissimilar to the anchor, respectively.
% 
For example, to learn the meaning of the concept ``book,'' the anchor might be an RGB image of a book; the corresponding text description and depth image of that book are included in the positive set; and the negative set includes the text description and RGB image of an apple.
% 
The methods we discuss here differ both in how they choose these three sets and the objective function they use.

@@

As originally proposed, the \supcon{} approach which we choose as our baseline was originally applied to unimodal datasets such as ImageNet~\cite{imagenet2009}, CIFAR-10~\cite{krizhevsky2009learning}, and CIFAR-100~\cite{krizhevsky2009learning}. We both demonstrate that it is possible to use on multimodal datasets and extend it with the addition of geometric approaches, which gives benefits on training speed and performance when modalities are dropped.

@@

Our proposed multimodal method is composed of two complementary parts. The first part is a geometric loss based on distances in the latent space, and the second part is a supervised contrastive loss based on cross-entropy (\supcon{}). The geometric loss is faster to learn while the cross-entropy based \supcon{} loss is more aligned with the downstream task of object retrieval. Hence we propose to combine them.

@@

To address this issue, we use pairwise distance optimization for all data points. %
Our method can be used for an arbitrary number of modalities. %
In contrast to our work, previous works that use triplet loss~\citep{GoLD_UMBC,triplet_loss_2021_CVPR} concatenate RGB and depth to form a single ``vision'' vector to handle three modalities, so they cannot robustly handle RGB or depth sensor ablation during test.
Moreover, our method has the advantage that it does not require providing hard negative examples.
Therefore, we modify the concept of triplet loss as follows. During training, we sample two different instances and their corresponding representations from all modalities into two sets---one positive set (referring to a specific object) and one negative set (referring to some different object) as shown in \cref{fig:emma-loss-trimmed}.
Unlike some prior triplet loss methods~\citep{GoLD_UMBC,triplet_loss_2021_CVPR}, the anchor is not randomly chosen from different modalities in each batch. Instead, in our setting, every item in the positive set becomes anchor once, and we minimize the distance between that item and other items in the positive set, while minimizing the distance between that item and all items in the negative set. It can be seen as an one-to-many relationship instead of an one-to-two relationship in the triplet loss formulation.
% the anchor and positive items are from the same object class, but in different modalities and each of the positives modalities becomes anchor once. 
To make the distinction between triplet loss and our proposed \geom{} more clear, we provide an example for the terminology we use.
% \todokdinline{I feel like using the terminology of triplet loss especially \textbf{anchor} is confusing here or maybe it needs to be rephrased. We can alternatively say that we consider all pairwise distances.}
% \todoffinline{I don't see the problem with anchor, but if you'd rather use a different term, what about query? As a note, while you do consider all pairwise distances but you treat some differently than others.}
\begin{itemize}
    % \item data point: a row in the csv/tsv file.
    \item Positive (Instance): A set of embeddings of one data point (e.g., an RGB image of an apple, corresponding depth image, text description, and speech description of the same apple) as shown with green rectangles in \cref{fig:emma-loss-trimmed}
    \item Negative (Instance): A set of embeddings of another data point of a different object (e.g., an RGB image of a mug, corresponding depth image, text description, and speech description of the same mug) as shown with orange rectangles in \cref{fig:emma-loss-trimmed}
    \item Anchor (Modality): Every modality of the positive set is chosen as the learning anchor once. In our formulation, anchor is a ``concept'' referring to the points in the positive sets rather than a single instance in itself. In \cref{fig:emma-loss-trimmed}, each of the 4 modalities is selected as an anchor once.
    %For our experiments, text is used. 
    The anchor is used as the basis for learning distances between positive and negative samples.
    % variance of text is lower and clusters of different objects are more distinct compared to other modalities, therefore, it can train a better model.
    % where we don't minimize the distance between the pair of embeddings from all modalities of the same data point, and we don't maximize the distance between positive and negative datapoints of the same modality.
    % \todo{explain this better}
    % \todokd{I rephrased it. Good?}
\end{itemize}

@@

\paragraph{Combining Geometric Loss and Cross-Entropy based \supcon{} Loss}
\label{subsec:emma}

@@

% 
% \begin{comment}
% 
% \begin{equation}\label{eq:emma}
%     % \mathcal{L} = \geom{} + \supcon{}
%     \ours{} = \geom{} + \supcon{}
% \end{equation}
% \end{comment}
% 
Both the geometric alignment and cross-entropy portions of EMMA are designed to not anchor on any specific modality, but instead consider all available modalities. This geometric portion is in contrast to previous triplet-based approaches. To show this, we consider how the geometric portion of EMMA’s loss would apply to the example in Figure 2. Considering the apple (on the left) as instance i, the dotted lines connecting the embeddings of the apple’s modalities reflect minimizing the pairwise distances between the different modalities of i: in equation 8, we do this via the $h(x, y)$ function. The dashed lines connecting the representations of the apple to a not-apple (the coffee mug) reflect maximizing the distances between these embeddings (done via $g(x,y)$). Notice how in both cases we consider all relevant pairs.

Although both \supcon{} and the \geom{} approaches are trying to bring target objects together (and push dissimilar ones away), there is a trade-off in \supcon{} where there is a normalized ranking across them. In contrast, in \geom{} distances can be arbitrarily near or far. Our approach tries to respect distinct modalities of each instance.
While our formulation is similar to \supcon{} notationally, conceptually \supcon{} uses different augmentations of RGB images as positive examples (inherently coming from the same underlying empirical distribution), while we use different modalities drawn from different distributions. To the best of our knowledge our work is the first to use supervised contrastive learning for such a multimodal scenario, where the different language and sensor inputs provide more robust learning than can be accomplished with single sensor inputs, even augmented.

Compared to pure \supcon{}, combining these loss functions results in faster convergence and slightly improved performance when all modalities are available, and maintains improved performance when modalities are missing. Experimental results are presented in detail in \cref{sec:results}.

@@

Our time analysis shows that EMMA takes almost 8 epochs to converge and each epoch takes roughly 0.7 minutes, which makes it 5.6 minutes until convergence, while \supcon{} takes about 36 epochs to converge and each epoch takes 0.52 minutes which amounts to 18.72 minutes. That is when we use all four modalities for training. When we ablate one or two modalities, the training takes less time.

@@

\subsection{Discussion}
Our proposed model performs well and learns fast, has been demonstrated to handle four modalities of shared information effectively, and is robust to test-time situations where information from one or more modalities is missing. 
The bottleneck for agents in different settings may differ, and training speed may not be critical in offline learning scenarios. However, since we usually need to finetune models for different tasks when it comes to transfer learning, the training speed becomes relevant.

There remains room for improvement. Specifically, the speech modality is harder to handle. \Cref{fig:3d-tsne} shows that although the relative position of instances are correct in the speech space, the distinction and clustering of different objects are not as good as the other three modalities.
